Adaptive Approximation for Multivariate Linear Problems \\ with Inputs Lying in a Cone

### 2018-11-09 RICAM2018NovTalk.tex
polish

\thanksnote{{\large Joint work with Yuhan Ding, Peter Kritzer, and Simon Mak} \\
	This work partially supported by  NSF-DMS-1522687 and NSF-DMS-1638521 (SAMSI) \& ???
}
\event{RICAM Workshop on Multivariate Algorithms and Information-Based Complexity}
\date[]{\\ November 9, 2018}

@@

\centerline{
    \begin{tabular}{>{\vspace{-1.7ex}\raggedleft}p{0.2\textwidth}@{\qquad}p{0.33\textwidth}@{\qquad}p{0.27\textwidth}}
    & \alert{Tidbits from talks this week} & \alert{My reponse} \tabularnewline
    \toprule
    \alert{Henryk, Klaus, Greg, Stefan, Erich, \ldots} & Many results for $f \in \smallscoop$  &
    Let's obtain analogous results for $f \in \smallcone$? \tabularnewline[2ex]
    \alert{Houman} & Let's \alert{learn} the appropriate kernel from the function data & Will only work for nice functions in a \smallcone \tabularnewline[2ex]
    \alert{Klaus} & ``This adaptive algorithm has \alert{no theory}''  & We want to construct adaptive algorithms \alert{with theory} \tabularnewline[2ex]
    \alert{Henryk} & Tractability & Yes! \tabularnewline[2ex]
    \alert{Greg} & POD weights & Yes! \tabularnewline[2ex]
    \alert{Mac} & Function values are expensive & Yes!
    \end{tabular}}
\end{frame}

@@

    \item \alert{Cone} means $f \in \cc \implies af \in \cc \ \forall a \in \reals$

@@

    \item If we can bound the error for $f \in $ \alert{cone}, then we can typically also bound the error  for $af$

@@

\alert{Definition} \quad $(S:\cc \subseteq \cf\to\cg,\Lambda)$ \alert{solvable} $\iff \ca(\cc) \ne \emptyset$\\[1ex]
\alert{Lemma}  \quad $f_1, f_2 \in \cc$  and $A(f_1, \varepsilon) = A(f_2,\varepsilon) \implies \norm[\cg]{S(f_1-f_2)} \le 2\varepsilon$\\[1ex]
\alert{Corollary} \label{ZeroCorollary} \quad $(S:\cc \subseteq \cf\to\cg,\Lambda)$ \alert{solvable} and $\exists f \in \cc,\  \varepsilon > 0$ with $A(f,\varepsilon) = A(0,\varepsilon)$  \\ \hfill \hfill $\implies S(f) = 0$\\[1ex]
\alert{Theorem} \label{VectorSpaceThm}\quad $(S:\cc \subseteq \cf\to\cg,\Lambda)$ \alert{solvable} and $\cc$ is a \alert{vector space} \\ \hfill \hfill 
$\iff S(f) = \sum_{i=1}^n L_i(f) g_i$ for some $\vL \in \Lambda^n, \ \vg \in \cg^n, \ n \in \naturals$  \hyperlink{VectorSpaceThmProof}{\beamergotobutton{Proof}}\\[1ex]
E.g. $\left(\int_{[0,1]^d} \cdot (\vx) \, \dif \vx: C^{1,\ldots,1}[0,1]^d\to\reals,\Lambda^{\textup{std/\alert{all}}}\right)$ is unsolvable/\alert{solvable}

@@

\vspace{-8ex}
\begin{gather*}
    \text{\alert{Input: }} f = \sum_{\vj \in \cn} \hf(\vj) u_{\vj} , \qquad 
    \text{\alert{Output: }} g = \sum_{\vj \in \cn} \hg(\vj) v_{\vj}, \qquad S(u_{\vj}) = v_{\vj}, \\
    \text{E.g., Function recovery: } S(u_{\vj}) = u_{\vj} = v_{\vj}, \qquad
   \text{ Integration: } S(u_{\vj}) = \int_{[0,1]^d} u_\vj (\vx) \, \dif \vx  = v_\vj \\
    \text{\alert{Fixed sample size algorithm: }} \Sapp(f,n) = \sum_{i=1}^n L_i(f) g_i   \\
    \alert{\app(f,\varepsilon) = \Sapp(f,n) + \text{ stopping criterion}}
\end{gather*}

@@

\vspace{-4ex}
\alert{Have} $\cost(f,A,\varepsilon)$; \quad \alert{No} $\cost(A,\cc,\varepsilon,\rho)$,  $\comp(\ca(\cc),\varepsilon,\rho)$, or  tractability results \alert{yet}

@@

\begin{align*}
    \only<1>{\cf &:= \left \{ f = \sum_{\vj \in \cn} \hf(\vj) u_{\vj} : \norm[\cf]{f} := \norm[2]{\left(\frac{\bigabs{\hf(\vj)}}{\lambda_{\vj}} \right)_{\vj \in \cn}} \right \} \qquad \begin{minipage}{3cm}\raggedright\alert{$\vlambda$ affects convergence rate \& tractability}\end{minipage}\\
    \cg &: = \biggl \{ g = \sum_{\vj \in \cn} \hg(\vj) v_{\vj} : \norm[\cg]{g} := \bignorm[2]{\hg}\biggr \}, \qquad v_{\vj} = S(u_{\vj}) \\[-2ex]}
      & \alert{\lambda_{\vj_1} \ge \lambda_{\vj_2} \ge \cdots}, \qquad
      n_0 < n_1 < n_2 < \cdots,  \qquad
 	\sigma_k(f) = \sum_{i=n_{k-1}+1}^{n_k}  \bigabs{\hf(\vj_i)}, \quad k \in \naturals \\
     \cc &: = \left\{f \in \cf :  
 	 \sigma_\ell(f) \le a b^{\ell - k}\sigma_k(f) \quad \forall k \le \ell \right\} \quad a > 1 > b \quad \text{\alert{series coef.\ decay steadily}}\\
 	 \Sapp(f,n_k) &= \sum_{i=1}^{n_k} \hf(\vj_i) v_{\vj_i} \text{ is optimal for fixed }n_k, \qquad \norm[\cg]{S(f) - \Sapp(f,n_k)} \le \frac{ab \sigma_k(f)}{\sqrt{1 - b^2}} \\[-1ex]
 	 \alert{A(f,\varepsilon)} & \alert{= \Sapp(f,n_k)} \text{ for the smallest $k$ satisfying } \alert{\sigma_k(f) \le \frac{\varepsilon \sqrt{1 - b^2}}{ab}}
 	 \only<2>{\\ 
 	 \MoveEqLeft{\alert{\cost(A,\cc,\varepsilon,\rho) = n_{\ell^\dagger}}, \quad 
 	 \ell^\dagger \le \min \left \{\ell \in \naturals : \frac{\rho^2}{\varepsilon^2} \le \frac{(1 - b^2)}{a^2b^2} \left[ \sum_{k=1}^{\ell-1} \frac{b^{2(k-\ell)}}{a^2\lambda_{n_{k-1}+1}^2} + \frac{1}{\lambda_{n_{\ell-1}+1}^2}\right]   \right\}} \\
 	 \MoveEqLeft{\cost(A,\cc,\varepsilon,\rho) \text{ \alert{essentially no worse than} } \comp(\ca(\cc,\LambdaAll),\varepsilon,\rho)}\\
 	 \MoveEqLeft{\text{\alert{No} tractability results}}}
\end{align*}

@@

    \begin{align*}
    \only<1>{\cf &:= \left \{ f = \sum_{\vj \in \natzero^d} \hf(\vj) u_{\vj} : \norm[\cf]{f} :=  \norm[\infty]{\left(\frac{\bigabs{\hf(\vj)}}{\lambda_{\vj}} \right)_{\vj \in \cn}} \right \} \qquad \begin{minipage}{3cm}\raggedright\alert{$\vlambda$ affects convergence rate \& tractability}\end{minipage}\\
    \cg &: = \Biggl \{ g = \sum_{\vj \in \natzero^d} \hg(\vj) v_{\vj} : \norm[\cg]{g} := \bignorm[1]{\hg}\Biggr \}, \qquad v_{\vj} = S(u_{\vj}) = u_{\vj} \\[-2ex]}
    \lambda_{\vj} & = \Gamma_{\norm[0]{\vj}} \prod_{\substack{\ell = 1 \\ j_{\ell} > 0}}^d \gamma_\ell s_{j_\ell} \quad
    \begin{cases}
\gamma_\ell = \text{coordinate importance} \\[-0.5ex]
\Gamma_r = \text{order size} \\[-0.5ex]
s_j =  \text{smoothness degree}
\end{cases}
\text{\alert{POSD weights}} \\[-2ex]
      & \alert{\lambda_{\vj_1} \ge \lambda_{\vj_2} \ge \cdots}, \qquad
      \Sapp(f,n) = \sum_{i=1}^{n} \hf(\vj_i) u_{\vj_i} ,  \qquad
 	\norm[\cg]{S(f) - \Sapp(f,n_k)} \le \norm[\cf]{f} \sum_{i = n+1} \lambda_{\vj_i} \\
     \cc &: = \left\{f \in \cf :  \text{ those functions for which } \norm[\cf]{f} \text{ can be inferred from a pilot sample} \right \}
\end{align*}

@@

\begin{frame}{Cheng and Sandu Function\footfullcite{VirLib17a}}
    \vspace{-9ex}
    \begin{gather*}
        \text{Order weights }\Gamma_k = 1, \qquad \text{Coordinate weights } \gamma_\ell \text{ \alert{inferred}}, \\
        \text{Smoothness weights } s_j \text{ \alert{inferred}}, \qquad \LambdaStd
    \end{gather*}

@@

\section{Summary}
\begin{frame}{Take Home Message}

@@

\begin{itemize}
    \item Assuming that the input functions lie in \alert{cones} \smallcone allow us to construct \alert{adaptive} algorithms

    \item Cone definitions reflect \alert{prior beliefs} or \alert{practical considerations}

>>>>

\thanksnote{{\large Joint work with Yuhan Ding, Peter Kritzer, and Simon Mak} \\
	This work partially supported by  NSF-DMS-1522687 and NSF-DMS-1638521 (SAMSI)
}
\event{RICAM Workshop on Multivariate Algorithms and Information-Based Complexity}
\date[]{\\ November 9, 2018}

@@

\centerline{
    \begin{tabular}{>{\vspace{-1.7ex}\raggedleft}p{0.2\textwidth}@{\qquad}p{0.33\textwidth}@{\qquad}p{0.27\textwidth}}
    & \alert{Tidbits from talks this week} & \alert{My reponse} \tabularnewline
    \toprule
    \alert{Henryk, Klaus, Greg, Stefan, Erich, \ldots} & Many results for $f \in \smallscoop$  &
    Let's obtain analogous results for $f \in \smallcone$? \tabularnewline[4ex]
    \alert{Houman} & Let's \alert{learn} the appropriate kernel from the function data & Will only work for nice functions in a \smallcone \tabularnewline[2ex]
    \alert{Klaus} & ``This adaptive algorithm has \alert{no theory}''  & We want to construct adaptive algorithms \alert{with theory} \tabularnewline[2ex]
    \alert{Henryk} & Tractability & Yes! \tabularnewline[2ex]
    \alert{Greg} & POD weights & Yes! \tabularnewline[2ex]
    \alert{Mac} & Function values are expensive & Yes!
    \end{tabular}}
\end{frame}

@@

    \item \alert{Cone} means $f \in \cc \implies af \in \cc \ \ \forall a \in \reals$

@@

    \item If we can bound the $\norm[\cg]{S(f) - \Sapp(f,n)}$ for $f \in $ \alert{cone}, then we can typically also bound the error  for $af$

@@

\alert{Definition} \quad $(S:\cc \subseteq \cf\to\cg,\Lambda)$ \alert{solvable} $\iff \ca(\cc,\Lambda) \ne \emptyset$\\[1ex]
\alert{Lemma}  \quad $f_1, f_2 \in \cc$  and $A(f_1, \varepsilon) = A(f_2,\varepsilon) \implies \norm[\cg]{S(f_1-f_2)} \le 2\varepsilon$\\[1ex]
\alert{Corollary} \label{ZeroCorollary} \quad $(S:\cc \subseteq \cf\to\cg,\Lambda)$ \alert{solvable} and $\exists f \in \cc,\  \varepsilon > 0$ with $A(f,\varepsilon) = A(0,\varepsilon)$  \\ \hfill \hfill $\implies S(f) = 0$\\[1ex]
\alert{Theorem} \label{VectorSpaceThm}\quad $(S:\cc \subseteq \cf\to\cg,\Lambda)$ \alert{solvable} and $\cc$ is a \alert{vector space} \\ \hfill \hfill 
$\iff \exists \ n \in \naturals, \ \vL \in \Lambda^n, \ \vg \in \cg^n$ such that $S(f) = \sum_{i=1}^n L_i(f) g_i \ \forall f \in \cc$ \alert{exactly}   \hyperlink{VectorSpaceThmProof}{\beamergotobutton{Proof}}\\[1ex]
E.g. $\left(\int_{[0,1]^d} \cdot (\vx) \, \dif \vx: C^{1,\ldots,1}[0,1]^d\to\reals,\Lambda^{\textup{std/\alert{all}}}\right)$ is unsolvable/\alert{solvable}

@@

\vspace{-8ex}
\begin{gather*}
    \text{\alert{Input: }} f = \sum_{\vj \in \cn} \hf(\vj) u_{\vj} , \qquad 
    \text{\alert{Output: }} g = \sum_{\vj \in \cn} \hg(\vj) v_{\vj}, \qquad S(u_{\vj}) = v_{\vj}, \\
    \text{E.g., Integration: } S(u_{\vj}) = \int_{[0,1]^d} u_\vj (\vx) \, \dif \vx  = v_\vj, \qquad  \text{Function recovery: } S(u_{\vj}) = u_{\vj} = v_{\vj}, \qquad\\
    \text{\alert{Fixed sample size algorithm: }} \Sapp(f,n) = \sum_{i=1}^n L_i(f) g_i   \\
    \alert{\app(f,\varepsilon) = \Sapp(f,n) + \text{ stopping criterion}}
\end{gather*}

@@

\vspace{-4ex}
\alert{Have} $\cost(f,A,\varepsilon)$; \quad \alert{No} $\cost(A,\cc,\varepsilon,\rho)$,  $\comp(\ca(\cc,\LambdaStd),\varepsilon,\rho)$, or  tractability results \alert{yet}

@@

\begin{align*}
    \only<1>{\cf &:= \left \{ f = \sum_{\vj \in \cn} \hf(\vj) u_{\vj} : \norm[\cf]{f} := \norm[2]{\left(\frac{\bigabs{\hf(\vj)}}{\lambda_{\vj}} \right)_{\vj \in \cn}} \right \} \qquad \begin{minipage}{3cm}\raggedright\alert{$\vlambda$ affects convergence rate \& tractability}\end{minipage}\\
    \cg &: = \biggl \{ g = \sum_{\vj \in \cn} \hg(\vj) v_{\vj} : \norm[\cg]{g} := \bignorm[2]{\hg}\biggr \}, \qquad v_{\vj} = S(u_{\vj}) \\[-2ex]}
      & \alert{\lambda_{\vj_1} \ge \lambda_{\vj_2} \ge \cdots}, \qquad
      n_0 < n_1 < n_2 < \cdots,  \qquad
 	\sigma_k(f) = \sqrt{\sum_{i=n_{k-1}+1}^{n_k}  \bigabs{\hf(\vj_i)}^2}, \quad k \in \naturals \\
     \cc &: = \left\{f \in \cf :  
 	 \sigma_\ell(f) \le a b^{\ell - k}\sigma_k(f) \quad \forall k \le \ell \right\} \quad a > 1 > b \quad \text{\alert{series coef.\ decay steadily}}\\
 	 \Sapp(f,n_k) &= \sum_{i=1}^{n_k} \hf(\vj_i) v_{\vj_i} \text{ is optimal for fixed }n_k, \qquad \norm[\cg]{S(f) - \Sapp(f,n_k)} \le \frac{ab \sigma_k(f)}{\sqrt{1 - b^2}} \\[-1ex]
 	 \alert{A(f,\varepsilon)} & \alert{= \Sapp(f,n_k)} \text{ for the smallest $k$ satisfying } \alert{\sigma_k(f) \le \frac{\varepsilon \sqrt{1 - b^2}}{ab}}
 	 \only<2>{\\ 
 	 \MoveEqLeft{\alert{\cost(A,\cc,\varepsilon,\rho) = n_{\ell^\dagger}}, \quad 
 	 \ell^\dagger \le \min \left \{\ell \in \naturals : \frac{\rho^2}{\varepsilon^2} \le \frac{(1 - b^2)}{a^2b^2} \left[ \sum_{k=1}^{\ell-1} \frac{b^{2(k-\ell)}}{a^2\lambda_{n_{k-1}+1}^2} + \frac{1}{\lambda_{n_{\ell-1}+1}^2}\right]   \right\}} \\
 	 \MoveEqLeft{\cost(A,\cc,\varepsilon,\rho) \text{ \alert{essentially no worse than} } \comp(\ca(\cc,\LambdaAll),\varepsilon,\rho)}\\
 	 \MoveEqLeft{\text{\alert{No} tractability results}}}
\end{align*}

@@

    \begin{align*}
    \only<1>{\cf &:= \left \{ f = \sum_{\vj \in \natzero^d} \hf(\vj) u_{\vj} : \norm[\cf]{f} :=  \norm[\infty]{\left(\frac{\bigabs{\hf(\vj)}}{\lambda_{\vj}} \right)_{\vj \in \cn}} \right \} \qquad \begin{minipage}{3cm}\raggedright\alert{$\vlambda$ affects convergence rate \& tractability}\end{minipage}\\
    \cg &: = \Biggl \{ g = \sum_{\vj \in \natzero^d} \hg(\vj) v_{\vj} : \norm[\cg]{g} := \bignorm[1]{\hg}\Biggr \}, \qquad v_{\vj} = S(u_{\vj}) = u_{\vj} \\[-2ex]}
    \lambda_{\vj} & = \Gamma_{\norm[0]{\vj}} \prod_{\substack{\ell = 1 \\ j_{\ell} > 0}}^d \gamma_\ell s_{j_\ell} \quad
    \begin{cases}
\gamma_\ell = \text{coordinate importance} \\[-0.5ex]
\Gamma_r = \text{order size} \\[-0.5ex]
s_j =  \text{smoothness degree}
\end{cases} \quad\begin{minipage}{5.5cm}\raggedright\alert{POSD weights reflect effect sparsity, effect hierarchy, effect heredity, and effect smoothness}\end{minipage}
 \\[-2ex]
      & \alert{\lambda_{\vj_1} \ge \lambda_{\vj_2} \ge \cdots}, \qquad
      \Sapp(f,n) = \sum_{i=1}^{n} \hf(\vj_i) u_{\vj_i} ,  \qquad
 	\norm[\cg]{S(f) - \Sapp(f,n_k)} \le \norm[\cf]{f} \sum_{i = n+1} \lambda_{\vj_i} \\
     \cc &: = \left\{f \in \cf :  \text{ those functions for which } \norm[\cf]{f} \text{ can be \alert{inferred} from a pilot sample} \right \}
\end{align*}

@@

\begin{frame}{Cheng and Sandu Function\footfullcite{VirLib17a}}
    \vspace{-9ex}
    \begin{gather*}
        \text{Chebyshev polynomials}, \qquad \text{Order weights }\Gamma_k = 1, \qquad \text{Coordinate weights } \gamma_\ell \text{ \alert{inferred}}, \\
        \text{Smoothness weights } s_j \text{ \alert{inferred}}, \qquad \LambdaStd
    \end{gather*}

@@

\section{Summary}
\begin{frame}{Take Home Messages}

@@

\begin{itemize}
    \item Assuming that the input functions lie in convex \alert{cones} \smallcone allow us to construct \alert{adaptive} algorithms

    \item Cone definitions reflect \alert{prior beliefs} and/or \alert{practical considerations}
